Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit

### 2016-08-17 paper/paper.tex
Near final edits

The above results indicate that (1) the mass density in the at geosynchronous distances is best correlated with $F10.7$ and (2) there exists a much weaker statistical relationship between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \cite{Takahashi2010} data set and attempt to identify and statistically characterize the relationship between geomagnetic activity with mass density at geosynchronous distances.  In addition, we attempt to identify solar wind processes that may drive the changes.

@@

The top panel of Figure~\ref{fig:ccplot} shows the long-term trends of log$_{10}(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{fig:ccplot} along with that for 1--day and 1--hour medians.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used combined measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites is 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29, respectively.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index below the threshold of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by the $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmapause earthward (\cite{LemaireEarthsPlasmasphere}) and the second is a possible change in density due to magnetospheric processes.

The second type of event is an increase in $\rho_{eq}$ above a threshold of 20~amu/cm$^3$.  It is expected that these increases will typically occur after period of quiet geomagnetic activity but may also be the result of increases in solar wind driven geomagnetic activity. In this case, the plasmapause boundary may move outward past geosynchronous orbit and the increase will be due to measurements being made inside the plasmapause.  A second possible reason for the increase could be due to magnetospheric processes.

@@

Our first analysis uses GOES-6 data only from the time interval 1989-1991, which corresponds with the interval used in the epoch analysis of \cite{Takahashi2010}. In this interval, there were 75 $D_{st}$ events where $D_{st}$ remained below the $-50$~nT threshold for at least 12 hours. Only events with an onset time between 06:00 and 12:00 MLT were considered. The first hour of a local minimum in $D_{st}$ following a threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after event onset. For multiple detected threshold crossings within 24 hours of each other, the first crossing was selected as the event because visual inspection showed the following crossings to typically be from brief downward deviations during the recovery period of a geomagnetic storm. To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across events on each epoch day.

@@

In the lower panel of Figure \ref{fig:DailyAveragedDstEvents}, it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6 (1983-1991) - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.  This trend is also seen for GOES 7 (1987-1991), is much less significant for GOES 5 (1983-1987), and much more significant for GOES 2 (1981-1983), correlating with the average value of $F_{10.7}$ during each respective coverage period.

To determine if the magnitude of the density medians in the top panel of Figure~\ref{fig:DailyAveragedDstEvents} on epoch days -1, 0, and 1 are statistically different from each other, we used a two-population bootstrap test.  A bootstrap sample of the medians for each epoch day was created by sampling the values used to compute the median on each epoch day with replacement.  The observed difference between the medians is shown in Table~\ref{fig:BootstrapDifferenceTable} along with the fraction of observations in 1000 bootstrap differences with a different sign than that observed.

If a statistically significant difference in the medians between two days existed, the fraction of bootstrap samples with a different sign is expected to be much smaller ($<$ 5\%).  In terms of a hypothesis test, we cannot reject the null hypothesis that the medians are the same with a confidence higher than 100 minus the percentage value shown in the table; the highest confidence level of 86\% is in the difference between day 0 and day -1.

@@

\begin{table}
\small
\centering
\input{../tables/DeltaBootstraps-case10.txt}
\caption{Results of test on of means of $\rho_{eq}$ shown in the top panel of Figure~\ref{fig:DailyAveragedDstEvents} between days of threshold crossing  near (day = 1 or -1) or on the day of a $D_{st}$ event (day = 0).}
\label{BootstrapDifferenceTable}
\end{table}

@@

To increase the number of events, we have removed the constraint that the events must occur in the MLT range from 06:00-12:00 and the constraint that $D_{st}$ remained below the threshold value for 12 hours or longer.  The result is shown in the bottom panel of Figure~\ref{fig:HourlyAveragedDstEvents}. The two spikes (one upwards after onset, one downwards before onset) were tested for significant deviation from two baselines, one 24-12 hours before onset, one 16-48 hours after onset. The probability of the left spike coming from the same distribution as the left baseline is 1.06\%, and 8.2\% from the right baseline, whereas the probability of the right spike coming from the left baseline is 1.89\% and 0.17\% from the right baseline.  This significant result is also maintained for GOES 2, 5, and 7. This indicates that there is a statistically significant peak in $\rho_{eq}$ within 6 hours of onset when hourly averages are considered.  This result is consistent with previous observations of large mass density increases in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

@@

The process for creating Figure~\ref{fig:BzBinnedDstEvents} was repeated except using the sort variable $F10.7$ instead of $B_z$, and the result is shown in Figure~\ref{fig:F107BinnedDstEvents}. The difference between the epoch curves is statistically significant at all times, as expected due to the high correlation between $\rho_{eq}$ and $F10.7$, but elevated $F10.7$ is associated with a peak in the $\rho_{eq}$ curve approximately 5 hours after onset. The values contributing to the peak were compared to a baseline of all values 16-48 hours after onset via a Wilcoxon rank-sum test and were found to have significantly different medians at the 99\% significance level.  This indicates the possibility that there must be a high base-line level of $\rho_{eq}$ in order for an increase in $\rho_{eq}$ to be observed for a $D_{st}$ event.

@@

An alternative type of event can be defined that corresponds to large increases in $\rho_{eq}$.  Figure~\ref{fig:HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\rho_{eq}$ above $20$~amu/cm$^2$ defines the start of the event. In finding $\rho_{eq}$ events, missing values were replaced with linearly interpolated values.  These events are associated with positive $B_z$ and low values of $D_{st}$.  Approximately 12 hours after onset, $\rho_{eq}$ slowly decays over 40 hours to near pre--onset levels.

@@

Figure~\ref{fig:BzBinned} shows the the events of Figure~\ref{fig:HourlyAveragedRhoEvents} when separated by $B_z$ in the same manner as Figure~\ref{fig:BzBinnedDstEvents}.  The largest difference occurs when the separation is performed after onset, with more positive $B_z$ leading to a larger peak in $\rho_{eq}$ than more negative $B_{z}$.  The rate of change of $\rho_{eq}$ also appears to be insenstive to $B_z$ for approximately 12 hours prior to onset.

@@

Figure~\ref{fig:F107Binned} shows how the epoch average curve of $\rho_{eq}$ in Figure~\ref{fig:BzBinned} depends on $F10.7$.  Statistically significant differences occur independent of if the separating factor the average of $F10.7$ before or after onset, as expected because the cadence of $F10.7$ is one day and its time scale of variation is longer than that of $\rho_{eq}$.  Events with high $F10.7$ are associated with higher $\rho_{eq}$ before and after event onset but similar peak values.  As was the case for $B_Z$, there is an approximately 12--hour time span of linear growth of $\rho_{eq}$ near the time of onset but because the initial starting $\rho_{eq}$ is higher for high $F10.7$, the associated growth rate is lower.

@@

On average, around the time of large geomagnetic events, there is a statistically weak enhancement of daily-averaged $\rho_{eq}$, but consideration of hourly average $\rho_{eq}$ shows enhancements over a 6-hour window that contains few events.  For the hourly-averaged measurements, the average of $B_{z}$ around onset time did not have a significant impact on $\rho_{eq}$ after onset.

It was shown that a key factor in determining if large $\rho_{eq}$ enhancements will be observed is elevated $F10.7$.  When geomagnetic events were separated according to if they occured when $F10.7$ was low or high, significant enhancements in $\rho_{eq}$ were only observed under high $F10.7$ conditions.

For $\rho_{eq}$, more postive $B_{z}$ after onset lead to higher values of subsequent $\rho_{eq}$, which indicates that a substantial portion of the events may due to the apparent mass refilling mechanism described by \cite{Denton2016}.  Without additional information about the location of the plasmapause, it may not be possible to separate out the conditions required for enhancement due to enhanced magnetospheric activity from this mechanism.

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{figures/F107MD27d-GOES6.eps}
\includegraphics[scale=0.40]{figures/ccplot-GOES6.eps}
\caption{Top: 27-day non-overlapping medians of $F_{10.7}$ and $log(\rho_{eq})$ from GOES 6. Bottom: Correlation between $log(\rho_{eq})$ and $F_{10.7}$ using medians in non-overlapping hour, day, and 27-day windows.}
\label{fig:ccplot}
\end{figure}
\clearpage

@@

\clearpage
\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedB_z-case1-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedB_z-case1-t025-tf30-GOES6.eps}
  \caption{$D_{st}$ events of Figure~\ref{fig:HourlyAveragedDstEvents} separated by the average $D_{st}$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:BzBinnedDstEvents}
\end{figure}

\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedF_{10.7}-case1-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedF_{10.7}-case1-t025-tf30-GOES6.eps}
  \caption{$D_{st}$ events of Figure~\ref{fig:HourlyAveragedDstEvents} separated by the average $F10.7$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:F107BinnedDstEvents}
\end{figure}

>>>>

The above results indicate that (1) the mass density in the at geosynchronous distances is best correlated with $F10.7$ and (2) there exists a much weaker statistical relationship between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \cite{Takahashi2010} data set and attempt to identify and statistically characterize the relationship between geomagnetic activity with mass density at geosynchronous distances.  In addition, we attempt to identify solar wind variables that may drive the changes.

@@

The top panel of Figure~\ref{fig:ccplot} shows the long-term trends of log$_{10}(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{fig:ccplot} along with that for 1--day and 1--hour medians.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used combined measurements from all satellites in the time interval of 1980 through 1991 using the constraint $06\mbox{:}00 \leq MLT \le 12\mbox{:}00$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites is 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29, respectively.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index below the threshold value of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by the $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmapause earthward (\cite{LemaireEarthsPlasmasphere}) and the second is a possible change in density due to magnetospheric processes.

The second type of event is an increase in $\rho_{eq}$ above a threshold of 20~amu/cm$^3$.  It is expected that some of these increases will typically occur after period of quiet geomagnetic activity.  In this case, the plasmapause boundary may move outward past geosynchronous orbit and the increase will be due to measurements being made inside the plasmapause.  A second possible reason for the increase could be due to the same mechanism associated with increases that occur during enhanced geomagnetic activity.

@@

Our first analysis uses GOES-6 data only from the time interval 1989-1991, which corresponds with the interval used in the epoch analysis of \cite{Takahashi2010}. In this interval, there were 75 $D_{st}$ events where $D_{st}$ remained below the $-50$~nT threshold for at least 12 hours. Only events with an onset time between 06:00 and 12:00 MLT were considered. The first hour of a local minimum in $D_{st}$ following a threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after onset. For multiple detected threshold crossings within 24 hours of each other, the first crossing was selected as the event onset because visual inspection showed the subsequent crossings to typically be from brief downward deviations during the recovery period of a geomagnetic storm. To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across the averaged event time series on each epoch day.

@@

In the lower panel of Figure \ref{fig:DailyAveragedDstEvents}, it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6 (1983-1991) - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.  This trend is also seen for GOES 7 (1987-1991), is much less significant for GOES 5 (1983-1987), and much more significant for GOES 2 (1981-1983).  We note here that this trend has a relationship with the average value of F10.7 during each respective coverage period, with the most significance found during the 1981-1983 time interval when F10.7 was consistently elevated.

To determine if the magnitude of the density medians in the top panel of Figure~\ref{fig:DailyAveragedDstEvents} on epoch days -1, 0, and 1 are statistically different from each other, we used a two-population bootstrap test.  A bootstrap sample of the medians for each epoch day was created by sampling the values used to compute the median on each epoch day with replacement.  The observed difference between the medians is shown in Table~\ref{BootstrapDifferenceTable} along with the fraction of observations in 1000 bootstrap differences with a different sign than that observed.

If a statistically significant difference in the medians between two days existed, the fraction of bootstrap samples with a different sign is expected to be much smaller ($<$ 5\%).  In terms of a hypothesis test, we cannot reject the null hypothesis that the medians are the same with a confidence higher than 100 minus the percentage value shown in the table; the highest confidence level of 86\% is in the difference between days -1 and 0.

@@

\begin{table}
\small
\centering
\input{../tables/DeltaBootstraps-case10.txt}
\caption{Results of test on of medians of $\rho_{eq}$ shown in the top panel of Figure~\ref{fig:DailyAveragedDstEvents} between days of threshold crossing  near (day = 1 or -1) or on the day of a $D_{st}$ event (day = 0).}
\label{BootstrapDifferenceTable}
\end{table}

@@

To increase the number of events, we have removed the constraint that the events must occur in the MLT range of 06:00-12:00 and the constraint that $D_{st}$ remained below the threshold value for 12 hours or longer.  The result is shown in the bottom panel of Figure~\ref{fig:HourlyAveragedDstEvents}. The two variations from a baseline (one downwards before onset and one upwards after onset) were tested for significant deviation from two baselines, one 24-12 hours before onset, one 16-48 hours after onset. The probability of the downward variation coming from the same distribution as the data from the left baseline is 1.06\%, whereas the probability of the right upward variation coming from the same distribution of the data for the right baseline is 0.17\%.  This significant result is also maintained for GOES 2, 5, and 7.  This indicates that there are statistically significant local variation in $\rho_{eq}$ within 6 hours of onset when hourly averages are considered.  The signficance of the upward variation result is consistent with previous observations of large mass density increases in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

@@

The process for creating Figure~\ref{fig:BzBinnedDstEvents} was repeated except using the sort variable $F10.7$ instead of $B_z$, and the result is shown in Figure~\ref{fig:F107BinnedDstEvents}. The difference between the epoch curves is statistically significant at all times, as expected due to the high correlation between $\rho_{eq}$ and $F10.7$, but elevated $F10.7$ is associated with a peak in the $\rho_{eq}$ curve approximately 5 hours after onset. The values contributing to the peak were compared to a baseline of all values 16-48 hours after onset via a Wilcoxon rank-sum test and were found to have significantly different medians at the 99\% confidence level.  This indicates the possibility that there must be a high pre-existing baseline level of $\rho_{eq}$ in order for an increase in $\rho_{eq}$ to be observed for a $D_{st}$ event.  This result is also consistent with the the analysis of Figure~3 when repeated using measurements from GOES satellites taken during different parts of the solar cycle; the significance of the differences between the days around onset were largest when the measurements were take during time intervals when F10.7 was consistently elevated.

@@

An alternative type of event can be defined that corresponds to large increases in $\rho_{eq}$.  Figure~\ref{fig:HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\rho_{eq}$ above $20$~amu/cm$^2$ defines the start of the event. In finding $\rho_{eq}$ events, missing values were replaced with linearly interpolated values.  These events are associated with on-average positive $B_z$ and small values of $D_{st}$ with little variability.  Approximately 12 hours after onset, $\rho_{eq}$ slowly decays over 36 hours to near pre--onset levels.

@@

Figure~\ref{fig:BzBinned} shows the the events of Figure~\ref{fig:HourlyAveragedRhoEvents} when separated by $B_z$ in the same manner as Figure~\ref{fig:BzBinnedDstEvents}.  The largest difference occurs when the separation is performed after onset, with more positive $B_z$ leading to a larger peak in $\rho_{eq}$ than more negative $B_{z}$.  The rate of change of $\rho_{eq}$ also appears to be insensitive to $B_z$ for in the time windows used to determine the separation.

@@

Figure~\ref{fig:F107Binned} shows how the epoch average curve of $\rho_{eq}$ in Figure~\ref{fig:HourlyAveragedRhoEvents} depends on $F10.7$.  Statistically significant differences occur independent of if the separating factor the average of $F10.7$ before or after onset, as expected because the cadence of $F10.7$ is one day and its time scale of variation is longer than that of $\rho_{eq}$.  Events with high $F10.7$ are associated with higher $\rho_{eq}$ before and after event onset but similar peak values.  As was the case for $B_z$, there is an approximately 12--hour time span of linear growth of $\rho_{eq}$ near the time of onset but because the initial starting $\rho_{eq}$ is higher for high $F10.7$, the associated growth rate is lower.

@@

On average, around the time of large geomagnetic events, there is a statistically weak enhancement of daily-averaged $\rho_{eq}$. For hourly averages $\rho_{eq}$ shows enhancements over a 6-hour window that contains too few events for a statistical analysis.  When the constraint of extended geomagnetic events (greater than 12 hours below a threshold value) and the MLT of the observation of $\rho{eq}$ (06:00-12:00) is removed to allow for more events, statistically significant variations are found, with a peak occurring approximately six hours after the event onset.  For the hourly-averaged measurements, the average of $B_{z}$ around onset time did not have a significant impact on $\rho_{eq}$ after onset.

It was shown that a key factor in determining if large $\rho_{eq}$ enhancements will be observed is elevated $F10.7$.  When geomagnetic events were separated according to if they occurred when $F10.7$ was low or high, significant enhancements in $\rho_{eq}$ above a baseline value were only observed under high $F10.7$ conditions.

For $\rho_{eq}$, more positive $B_{z}$ after onset lead to higher values of subsequent $\rho_{eq}$, which indicates that a substantial portion of the events may due to the apparent mass refilling mechanism described by \cite{Denton2016}.  Without additional information about the location of the plasmapause, it may not be possible to separate out the conditions required for enhancement due to enhanced magnetospheric activity from this mechanism.

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{figures/F107MD27d-GOES6.eps}
\includegraphics[scale=0.40]{figures/ccplot-GOES6.eps}
\caption{Top: 27-day non-overlapping medians of $F_{10.7}$ and $\log_{10}(\rho_{eq})$ from GOES 6. Bottom: Correlation between $\log_{10}(\rho_{eq})$ and $F_{10.7}$ using medians in non-overlapping hour, day, and 27-day windows.}
\label{fig:ccplot}
\end{figure}
\clearpage

@@

\clearpage
\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedB_z-case1-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedB_z-case1-t025-tf30-GOES6.eps}
  \caption{$D_{st}$ events of the bottom panel of Figure~\ref{fig:HourlyAveragedDstEvents} separated by the average $D_{st}$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:BzBinnedDstEvents}
\end{figure}

\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedF_{10.7}-case1-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedF_{10.7}-case1-t025-tf30-GOES6.eps}
  \caption{$D_{st}$ events of the bottom panel of Figure~\ref{fig:HourlyAveragedDstEvents} separated by the average $F10.7$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:F107BinnedDstEvents}
\end{figure}
